---
title: "Simple Plotting in R"
author: "Victor Ordu"
date: "11 June 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

- `base` (Not actually a package)
- `ggplot2`


## Our Dataset
We will work with the in-built `airquality` dataset, which represents daily air quality measurements in New York from 1st May to 30th September 1973
```{r data}
df <- airquality
str(df)
```

## Matrix of scatterplots

```{r pairs}
pairs(df)
```

## Plotting of Numerical data
Histograms are particulary great for representing **discrete** numerical data i.e. integers.
```{r}
hist(df$Ozone)
```

## Plotting of Numerical Data (II)
Histograms are good at displaying the 'skew' of the distribution around an estimator such as the arithmetic mean or the median.

```{r hist2}
## Compare this with the other histogram
hist(df$Solar.R)
```

## Frequency polygon
Like histograms, frequency polygons are good for discrete values.

For Ozone:
```{r}
suppressPackageStartupMessages(library(mosaic))
freqpolygon(df$Ozone)
```


## Frequency Polygons (II)

For Solar Radiation:
```{r}
freqpolygon(df$Solar.R)
```
## Scatterplot

```{r scatter}
plot(df$Ozone)
```


## Do some data transformation

```{r transform}
df$Mth.Name <- month.name[df$Month]
df$Mth.Name <-
  factor(df$Mth.Name, levels = month.name[5:9], ordered = TRUE)
head(df$Mth.Name)
```

```{r month tail}
tail(df$Mth.Name)
```

We have created a new column called `Mth.Name` and made it an ordered factor (categorical variable) of the months of May through September. Can be confirmed using functions like `str`, `dim`, etc.

## Bar charts

```{r bar chart}
tab <- table(df$Mth.Name)
barplot(tab)
```


## Box-and-whiskers plot

```{r box-and-whiskers}
boxplot(df$Ozone)
```


## 1-D Scatter plot
```{r strip 1-D scatter}
stripchart(df$Ozone)
```


## Time Series

```{r time series}
ozts <-
  ts(df$Temp, start = 5, end = 9, frequency = length(levels(df$Mth.Name)))
plot(ozts)

```

